---
title: "Peer Assessment 1"
author: "rdh927"
date: "Monday, October 12, 2015"
output: html_document
---
## Loading and preprocessing the data
```{r, echo = FALSE}
## Set the working directory
setwd("C:/Users/Rachel/Documents/R Programming/reproddata")
```
```{r, echo=TRUE}
## Read in the dataset
activity <- read.csv("activity.csv")

## Calculate the total number of steps taken per day
total_steps <- tapply(activity$steps, activity$date,sum, na.rm=TRUE)
```
No additional processing/transformation was done prior to analysis.

## What is mean total number of steps taken per day?
```{r, echo=TRUE}
## Make a histogram of the total steps per day
hist(total_steps)

## Calculate the mean total steps/day
mean_steps <- mean(total_steps)

## Calculate the median total steps/day
median_steps <- median(total_steps)
```
The mean total number of steps taken per day is 9354. The median is 10,395.

## What is the average daily activity pattern?
```{r, echo=FALSE}
mean_steps_int <- tapply(activity$steps, activity$interval,mean, na.rm=TRUE)
plot(activity$interval[1:288],mean_steps_int, type="l",ylab = "Steps", xlab = "Interval (min)")
```
```{r, echo=TRUE}
## Interval containing maximum number of steps?
max(mean_steps_int)
print(subset(mean_steps_int, mean_steps_int >= 206))
```
The inverval containing the maximum number of steps is 835-840 minutes.

## Inputting missing values
The original dataset contained missing values; depending on how these missing values are treated, they may affect the results of our analysis. We replaced the missing values with the mean number of steps averaged over all days and intervals, and performed our analyses again.

```{r, echo=TRUE}
## How many NAs are there?
nona <- na.omit(activity)
num_nas <- as.numeric(nrow(activity)-nrow(nona))

## Fill in missing values with average number of steps for a given interval
## create new data set with missing values filled in
activity2 <- activity
activity2$steps[which(is.na(activity2$steps))] <- mean(mean_steps_int)
```
The number of NAs is 2304.

```{r, echo=FALSE}
## make a histogram of the total number of steps taken each day; report the mean and median total number of steps taken per day.
## compare to ignoring NAs. what is the impact of inputting missing data?
total_steps2 <- tapply(activity2$steps, activity2$date,sum, na.rm=TRUE)
hist2 <- hist(total_steps2)
```
```{r, echo=TRUE}
## Calculate the mean total steps/day
mean_steps2 <- mean(total_steps2)

## Calculate the median total steps/day
median_steps2 <- median(total_steps2)
```
Upon replacement of the NA values for the average number of steps, we find that the new mean and median are both 10766. The median has increased slightly, while the mean has increased considerably and become much closer to the median, meaning the data has, as we expect, become much more well-distributed (more normal).

## Are there differences in activity patterns between weekdays and weekends?
```{r, echo=TRUE}
## Create new factor with two levels - "weekday" and "weekend" (use weekdays() function)
activity$date <- as.character(activity$date)
activity$date <- as.POSIXct(activity$date, format="%Y-%m-%d")

weekdays1 <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')
## create a logical vector with `%in%` and `weekdays` 
## convert to factor and specify the `levels`
activity$weekday <- factor((weekdays(activity$date) %in% weekdays1),levels=c(FALSE, TRUE), labels=c('weekend', 'weekday'))

## subset activity into weekend and weekday days
weekdays <- subset(activity, weekday == 'weekday')
weekends <- subset(activity, weekday == 'weekend')

mean_steps_weekend <- tapply(weekends$steps, weekends$interval,mean, na.rm=TRUE)
mean_steps_weekday <- tapply(weekdays$steps, weekdays$interval,mean, na.rm=TRUE)
```

```{r, echo=FALSE}
## Two-panel plot comparing weekdays and weekends
par(mfrow=c(2,1))
plot(activity$interval[1:288],mean_steps_weekend, type="l", main = "weekend",ylab = "Steps", xlab = "Interval (min)")
plot(activity$interval[1:288],mean_steps_weekday, type = "l", main = "weekday",ylab = "Steps", xlab = "Interval (min)")
```

